Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2014-06-02 pituba/article/article_spe.tex
last hotel

The term black-oil refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined for a simplified description gas phase composition. A pseudo oil component can also be defined by subtracting the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from usually a large number to only two.

@@

Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase may be composed uniquely by gas component and, water phase, uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases.

@@

Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an agelbraic equation that can be solved to any of its terms provided all remaining are known. This works focuses on its solution for pressure.

@@

Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure. In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, and they are function of pressure. It turns out, though, that pressure is a function of depth and phase.

@@

Above relations, specially Eq.~\eqref{eq: PVTH1}, provide a test to check whether PVT data obtained from laboratory or empirical correlations is physically correct.

@@

\subsection{Solution algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces an artificial smoothness to the gradient direction near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.

>>>>

The term black-oil in this context refers to specific hydrocarbon compositions to which phase equilibrium for isothermal changes in pressure can be described by a two pseudo-component system. For such reservoir fluids, gas phase composition remains essentially constant in the range of pressures observed during production and a pseudo gas component can be defined. Similarly, a pseudo oil component can also be defined by subtracting all the pseudo gas component from oil phase composition. After defining these two pseudo-components, the number of transport equations needed for hydrocarbon species is reduced from a usually large number to only two.

@@

Reservoir fluids are composed of oil and gas pseudo-components and water component. There can be three fluid phases: oil, gas and (liquid) water. Oil phase may be composed of oil and gas components, gas phase can be composed uniquely by gas component and water phase uniquely by water component. The only allowed mass transfer is of gas component between oil and gas phases, i.e., composition of gas phase is always that of a dry gas.

@@

Eq~\eqref{eq: MBE} involves fluid properties, original fluid volumes ($N$, $W$ and $G)$, cumulative production and injection and aquifer influx.
It is an agelbraic equation that can be solved to any one of its terms, provided that all remaining are known. This works focuses on its solution for pressure.

@@

Material balance equation is widely used to stablish a relationship between production (and injection) and reservoir pressure. In Eq.~\eqref{eq: MBE}, however, formation volume factors were defined using averaged values of fluid properties, which are functions of local pressure. It turns out, though, that pressure is a function of depth and phase.

@@

Eq.~\eqref{eq: PVTH1} also provides one condition to check whether PVT data obtained from laboratory or empirical correlations is physically correct. It can be alternatively formulated as shown below and indicates that any vaporization of oil phase is followed by an increase in its density.
\begin{equation}
\frac{1}{\rho_o}\frac{d\rho_o}{dY_o} \geq Y_o \left( \frac{\rho_o}{\rho_g} - 1 \right) \, .
\end{equation}

@@

\subsection{Solution algorithm}
Several methods exist to find roots of nonlinear continuous functions \citep{hamming1987numerical}. One possible choice is Newton's method, which linearizes the function using either its gradient or a secant line. The secant line approach introduces an numerical smoothness to the gradient near continuous but non-differentiable points, for what it is the approach used here to deal with the bubble-point pressure.
